Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-01-09 MScDiplom_Lemenkova.tex
remove leftover
added tables capturing and updated figures referencing

\section[Overview of literature]{Seagrass monitoring: overview of literature and research resources}
%\markright{II. Seagrass monitoring: overview of literature and research resources}
%\renewcommand{\headrulewidth}{0.4pt}

@@

Another well-known multispectral sensor, SPOT provides multispectral imagery with a spatial
resolution of 10 m, covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \href{http://www.spotimage.fr/}{SPOT} imagery was used for mapping beds of Posidonia oceanica in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}. The \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, offering multispectral and panchromatic imagery, was the first to collect publicly
available high-resolution imagery at 1- and 4-meter resolution from \href{http://www.geoeye.com}{Geoeye}. IKONOS
imagery has been applied for the seagrass mapping due to its high resolution and accessibility. Thus,
the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, Landsat TM, and CASI, show that in the blue part of the
spectrum, the best results are achieved by IKONOS and CASI, while Landsat TM has not high
enough resolution. It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that CASI enable to receive still more accurate results of the the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.
Another comparative analysis of the application of \href{http://www.itres.com/products/imagers/casi550}{CASI}, Landsat and \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird} imagery \cite{Phinn08}\label{Phinn08}
 demonstrates the high suitability of CASI images for the fine-scale mapping of seagrass
landscapes. Thus, CASI and Quickbird-2 images enable to identify even separate seagrass species
with small width and heterogeneous nature of the seagrass patches, which could not be detected using
Landsat TM images with their 30*30m resolution.
Advanced Spectrometer for Thermal Emission and Reflection Radiometer (\href{http://asterweb.jpl.nasa.gov/}{ASTER}), launched in
1999 onboard Terra sensor, provides high-resolution images of the Earth in 15 different bands of the
electromagnetic spectrum, ranging from visible to thermal infrared light  (Fig.\ref{fig:13}). It has diverse
subsystems for the visible near infrared (VNIR) with 15 m resolution, shortwave infrared (SWIR), and
thermal infrared (TIR) wavelength regions \cite{Kalinowski04}\label{Kalinowski04}. For each channel there is separate
onboard calibration (OBC) system, telescope with independent pointing and different detector technsology \cite{Arai05}\label{Arai05}. 
This makes ASTER imagery expecially suitable for the application towards detailed
mapping of surface temperature, emissivity and reflectance of objects and bathymetric elevations as
well. Successful application of the ASTER imagery towards mapping of submerged vegetation
reported, for example, by Hirose \cite{Hirose04} \label{Hirose04}. Other multispectra images have also been used for the
seagrass interpretation. For example, the success of using imagery from the multispectral airborne
scanner Daedalus AADS1268 is reported by Heege et al \cite{Heege03}\label{Heege03} where they aim at classification of
macrophytes in shallow waters of the Lake Constance.
In regard to the methods chosen for the image interpretation, the supervised classifications proves to
be the most worthy in a majority of case studies \cite{Palandro03,Peneva08}\label{Palandro03} \label{Peneva08}. While
often methods of the unsupervised classification are used as a tool for classifying submerged object
and features on the multispectral images and aerial imagery \cite{Fletcher09}\label{Fletcher09}, it still does not
provide common algorithms that can be applied to other images and regions. Thus, the strength of the
optical signal coming from the various types of the seafloor is strongly influenced by the effects of the
water column, its depth and chemical properties. Yet methods of the unsupervised classification
overleap mixed spectral effects of the water column which shift the real values of the spectra, with the
pure reflectance from the benthos; as a result, it may cause significant errors \cite{Dierssen03}\label{Dierssen03}.
 Therefore, for the accurate assessments of various seafloor cover types, water
column depth and its optical properties, methods of supervised classification should be preferably
used as being more suitable in classification and interpretations of imagery.
While this and other studies \cite{Phinn08}\label{Phinn08}demonstrated the advantages and success of the
application of multispectral imagery for the spectral discrimination of the seafloor cover types and
mapping the submerged landscapes on the basis of pixels classification, the application of data from
hyperspectral sensors has better potential due to their higher resolution.

\subsubsection{Review of hyperspectral imagery used for seagrass mapping}
The application of the hyperspectral sensors is most effective and provide more accurate classification
results (Fig.2.1.),

@@

due to their higher spectral resolution \cite{Bharathi03}\label{Bharathi03} with interval
narrows to 10 nanometers, while broadband sensors are limited to the spectral width of ca
150 nm (Fig.\ref{fig:14}). 
Hyperspectral imagery is acquired through the simultaneous acquisition of images in many
narrow, contiguous spectral bands from hyperspectral scanners (mostly) cover the 400-
to 2500-nm spectral bands \cite{Schmidt03}\label{Schmidt03}.
Perhaps, the most advantageous and general characteristics of hyperspectral imagery is its
high spectral resolution, desirable for the case of seagrass monitoring.

@@

The most suitable
scanner fit for detailed seagrass mapping would cover bands of 550-750 nm and have a spectrum
resolution of 5-15 nm \cite{Fyfe04}\label{Fyfe04}, which exactly characterizes typical airborne and
hyperspectral satellite scanners. Important features may be detected in the narrow
wavelengths of hyperspectral imagery, while this information can be lost in the broader
wavelengths of other sensors. With its 126 spectral bands, HyMap imagery enables to
distinguish features of interest, i.e. seagrass types \cite{Peneva07}\label{Peneva07}, which is the major
advantage of the of hyperspectral data for mapping landscapes of the seagrasses.
While comparing multispectral imagery with airborne hyperspectral, the last one showed higher
overall accuracies \cite{Phinn08}\label{Phinn08}. Several, regular and narrow (10 nm) spectral bands, specific for
hyperspectral imagery, are strong tool enabling to detect even slight and subtle differences in the
spectral reflectances between various seafloor types, e.g. different seagrass species at diverse depths,
algae, corals, dark-coloured sands or other types of sediments \cite{Hochberg03a}\label{Hochberg03a}. 
Therefore, there is great potential of the application of hyperspectral remote sensing imagery towards seagrass
mapping at species level, as long as they are distinguishable spectrally (Fig.\ref{fig:15}) which, for example, has been
tested in the case study of Australian marine ecosystems by Fyfe \cite{Fyfe04}\label{Fyfe04}. Application of various
classifications methods, inter alia maximum likelihood, minimum distance and means, towards
hyperspectral imagery \cite{Peneva08}\label{Peneva08}, combined with fieldwork measurements ensures accurate
mapping results with the maximum likelihood methos producing the best results.
Therefore, accurate mapping of the seagrass landscapes and other seafloor types using remote sensing
approaches requires application of high-spatial resolution (higher than 5 m) or hyperspectral imagery.
Comparative analysis of the application of hyperspectral and multispectral imagery towards the
seafloor types classification \cite{Hochberg03a}\label{Hochberg03a} demonstrates that coral, seagrasses and
sand wery vell distinguishable in their spectra with an overall classification accuracy of 98 percent.
However, the use of data from various sensors, both hyperspectral and multispectral, are possible and
reasonable, as soon as it meets the research specific objective. 
Thus, the use of multispectral imagery
with high spatial resolution is preferable to using hyperspectral medium resolution data in case of
mapping benthic vegetation in areas where the spatial heterogeneity is very high \cite{Vahtmae07}\label{Vahtmae07}.

@@

The research is technically based on WASI, for the analysis of spectral signatures, \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS} and \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine} software for the processing and analysing of aerial and broadband images, Google Earth and  \href{http://landsat.gsfc.nasa.gov/}{Landsat}.
\href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0} software is used for the spatial analysis, general mapping and cartographic layout
presentation; the raster processing techniques are applied for the detection of seagrass spatial
distribution using supervised classification.
The research data include Google Earth aerial images and scenes from the \href{http://landsat.gsfc.nasa.gov/}{Landsat TM and ETM+}
covering research period of 10 years in the same year time, taken from \href{http://www.usgs.gov/pubprod/}{USGS}, \href{http://glovis.usgs.gov/}{GloVis}, \href{http://www.osdpd.noaa.gov/ml/index.html}{NOAA} and \href{http://glcfapp.glcf.umd.edu:8080/esdi/index.jsp}{The Earth Science Data Interface}. 
The satellite imagery provides
vital information of the most recent changes in \textit{P.oceanica} within the coastal areas, as well as the
condition (poor or destroyed).
The raster processing includes making mosaic-like covering for the whole research area. The Google
Earth images are most appropriate for the detailed mapping than stellite than \href{http://landsat.gsfc.nasa.gov/}{Landsat} or \href{http://asterweb.jpl.nasa.gov/}{ASTER} enabling to produce accurate maps with correct results. Therefore, the main set of images for the
current work is Google Earth aerial images. The  \href{http://landsat.gsfc.nasa.gov/}{Landsat} images are used for the general overview.
Besides aerial and satellite photographs, data acquired during the fieldwork (21 Sep – 11 Oct) are
necessary addition to the mapping helping to solve problems of interpretation during the images
classification \cite{Pasqualini98a}\label{Pasqualini98a}. Therefore, this work includes sampling of the in-situ
measurements of the seagrass distribution.
The sampling stations were located in two candidate places on the northern (Ligaria) and southern
(Xerocampus) parts of Crete island, as these regions are well suitable for the seagrass, due to the
annual mean water temperatures and geological factors, i.e. seafloor conditions and sediments. The
field campaign has been carried out during the September-October period 2010. 
The information about the location of the seagrass (mostly represented by the P.oceanica species) is useful for the
understanding of the relationship between the spatial distribution of the seagrass and the environment
of the selected areas of Cretan beaches. The results of the videographic measurements are used for the
seafloor types detection, because the objects represented on the photos can be well distinguished and
classified according to the following well-known characteristics \cite{Butler87}\label{Butler87}: size (yet some
measurements are necessary for the similar-looking objects); shape (the general form is most reliable
evidence for identification); colour (common and reliable object indicator); texture (when changes in
tone are too small to be distinguishable, texture may assist identification, e.g., stippled, granular,
rough, smooth, etc.); associated features (those usually found near other objects, e.g., rocks and soil).

@@

The Ligaria Beach is a narrow, sandy and pebble beach (Fig.12), located ca 15 km north-west from Heraklion.
The in-situ measurements were conducted during the fieldwork in the period 21.09.2010-11.10.2010.
The fieldwork on seagrass monitoring included visual estimations and photo- and video footage of the
above-ground seagrass patches, sediment seafloor cover types, species compositions, water depth and
geograpic locations recorded using GPS.

@@

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras Olympus ST 8000 (Fig.\ref{fig:20}), at
five minutes interval along the tracklog path (Fig.\ref{fig:21}).
research sampling design also included measurements of depths, because bathymetry is one of the
most determinating factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, iPAQs, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The GPS allowed to capture measurement locations on the iPAQ,
encapsulated in a plastic waterproof Otterbox. Camera were adjusted horizontally by a leveller and
were mounted under the bottom of the boat to capture video footage and imagery (Fig.\ref{app:20}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. The
locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 m2 size each.
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects.

@@

Along the Ligaria beach there are several types of the seagrass landscapes, namely, segrass meadows
continuously covering vast areas (Fig.\ref{app:18}), aggregated seagrass patches, represented by separate
mattes with short irregular channels between them (Fig.\ref{app:18}) and isolated seagrass patches, or mattes,
which located separately from broad meadows (Fig.\ref{app:18}).
The results of the underwater videometric measurements of the Olympus cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass P.oceanica meadows
and nature of the seafloor cover types (rocks, sandy, mixed). Seagrass species on the Ligaria and Agia
Pelagia beach consist of P.oceanica. The types of sediment on the Ligaria vary from coarse sand,
P.oceanica patches, sand, mud, rock gravel and fine sand (Fig.\ref{app:20}).

@@

\subsection{Review of the collected data}
The remote sensing imagery were collected for the Crete island area covering the research period and
area, enabling the field data to be used for the calibration and validation. The available aerial and
satellite imagery are commonly used for the mapping of seagrass landscapes and their application is
proven by various research papers. The imagery include satellite multi-spectral imagery (Landsat-TM,
ETM+) and aerial imagery from the Google Earth.
The overview of the collected data enables to summarize their following types:
1. Spectra of P.oceanica and other seafloor types at different environmental conditions
2. Aerial imagery from the Google Earth
3. Satellite images from the open sources (Landsat)
4. Underwater videometric measurements of the Olympus cameras made during the ship route
The available satellite imagery (Fig.\ref{app:20}) and aerial images are read into the ArcGIS project.
The available broadband and hyperspectral remote sensing data are used for the mapping of the
seagrass in shelf areas no deeper than 4.0 meters, for the environmental monitoring in order to detect
the dynamics in changes of the seagrass distribution along Crete during the past 10 years using
different satellite images for 2000-2010.
The spatial resolution of Landsat ETM+ image is 30 m in the visible and near infrared bands (bands
1-5 and 7); the spatial resolution of ASTER 15 m for the visible and near-infrared bands. IKONOS
acquires data in 3 visible channels and NIR, with spatial resolution of 1-4 meters. ASTER and
IKONOS images are suggested to be included as soon as available, in addition to the Landsat. The
research includes fieldwork in-situ measurements of the seagrass distribution along the northern and
south-eastern coasts of Crete in chosen locations.
The optical measurements of the irradiance and radiance of the sea water and bottom cover types of
the seafloor have been received by the means of the optical sensors Trios-RAMSES Hyperspectral
UVA/UVB Irradiance Sensor and RAMSES-ARC Hyperspectral UV-VIS Radiance Sensor (Fig.2.3
and 2.4), both adjusted for the measurements of the irradiance and radiance.
The highest measured values are located in the diapason of 410-730 nm for the water irradiance and 430-650 for the water radiance (Fig.\ref{fig:23}).

%\begin{figure}[h]
	%\centering
	%\subfloat [Irradiance]{\label{fig:23a}\includegraphics[width=0.8\textwidth]{Fig-23a.jpg}}\\
	%\subfloat [Radiance]{\label{fig:23b}\includegraphics[width=0.8\textwidth]{G-Radiance_Bezier.jpg}}
	%\caption{Optical properties of the sea water with sediments}
	%\label{fig:23}
%\end{figure}

Afterwards, the measured values of the
radiance and irradiance, respectively, have been used for the computation of the spectral reflectance
properties of the sea water and bottom cover types (Fig.\ref{fig:23}). The spectral range of radiance
cover diapason of 320-950 nm, and ittadiance measurements are covered in the interval of 280-500
nm, which is suitable for a characteristics of seagrass reflectance.
Different curves on the radiance and irradiance graphs (Fig.\ref{fig:23}) represent several series of
the measurements. The values of the spectral reflectance are received from the
computations of these values using Excel mathematical formulae. The graphs shown on Fig.\ref{fig:23} display values of the radiance and irradiance of the sea water in Agia Pelagia, with and
without sediments and suspended particles, respectively.
\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{GNU-17.jpg}
\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete}
\label{fig:25}
\end{figure}

\subsection{Data pre-processing}
A collection of visible spectra of two seafloor cover types - \textit{P.oceanica} and carbonatic sand - were
tested for the spectral variability and separability under varying conditions of dfifferent environmental
constituents (e.g. depth, water content, sun angle etc), in order to determine the
potential that it may have on the approaches for further images processing and classification.The raw
initial data of measurements of the spectral reflectance have been preprocessed and analysed. These data
have been measured using different step of the wavebands: some measurements
were made with 3 nm interval, while others – using 4 and 1 nm step.

\begin{wrapfigure}{r}{0.6\textwidth}
\centering
\includegraphics[scale=0.23]{Fig-24.jpg}
\caption{Script written on Python, for the interpolation raw data of the Trios-RAMSES measurements}
\label{fig:24}
\end{wrapfigure}

>>>>

\section[Overview of literature]{Seagrass monitoring: overview of literature and research resources} \label{sec:2}
%\markright{II. Seagrass monitoring: overview of literature and research resources}
%\renewcommand{\headrulewidth}{0.4pt}

@@

Another well-known multispectral sensor, SPOT provides multispectral imagery with a spatial
resolution of 10 m, covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \href{http://www.spotimage.fr/}{SPOT} imagery was used for mapping beds of Posidonia oceanica in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}. The \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, offering multispectral and panchromatic imagery, was the first to collect publicly
available high-resolution imagery at 1- and 4-meter resolution from \href{http://www.geoeye.com}{Geoeye}. IKONOS
imagery has been applied for the seagrass mapping due to its high resolution and accessibility. Thus,
the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, Landsat TM, and CASI, show that in the blue part of the
spectrum, the best results are achieved by IKONOS and CASI, while Landsat TM has not high
enough resolution. It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that CASI enable to receive still more accurate results of the the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.
Another comparative analysis of the application of \href{http://www.itres.com/products/imagers/casi550}{CASI}, Landsat and \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird} imagery \cite{Phinn08}\label{Phinn08}
 demonstrates the high suitability of CASI images for the fine-scale mapping of seagrass
landscapes. Thus, CASI and Quickbird-2 images enable to identify even separate seagrass species
with small width and heterogeneous nature of the seagrass patches, which could not be detected using
Landsat TM images with their 30*30m resolution.

@@

Advanced Spectrometer for Thermal Emission and Reflection Radiometer (\href{http://asterweb.jpl.nasa.gov/}{ASTER}), launched in
1999 onboard Terra sensor, provides high-resolution images of the Earth in 15 different bands of the
electromagnetic spectrum, ranging from visible to thermal infrared light  (Fig.\ref{fig:13}). It has diverse
subsystems for the visible near infrared (VNIR) with 15 m resolution, shortwave infrared (SWIR), and
thermal infrared (TIR) wavelength regions \cite{Kalinowski04}\label{Kalinowski04}. For each channel there is separate
onboard calibration (OBC) system, telescope with independent pointing and different detector technsology \cite{Arai05}\label{Arai05}. 
This makes ASTER imagery expecially suitable for the application towards detailed
mapping of surface temperature, emissivity and reflectance of objects and bathymetric elevations as
well. Successful application of the ASTER imagery towards mapping of submerged vegetation
reported, for example, by Hirose \cite{Hirose04} \label{Hirose04}. Other multispectra images have also been used for the
seagrass interpretation. For example, the success of using imagery from the multispectral airborne
scanner Daedalus AADS1268 is reported by Heege et al \cite{Heege03}\label{Heege03} where they aim at classification of
macrophytes in shallow waters of the Lake Constance.
In regard to the methods chosen for the image interpretation, the supervised classifications proves to
be the most worthy in a majority of case studies \cite{Palandro03,Peneva08}\label{Palandro03} \label{Peneva08}. While
often methods of the unsupervised classification are used as a tool for classifying submerged object
and features on the multispectral images and aerial imagery \cite{Fletcher09}\label{Fletcher09}, it still does not
provide common algorithms that can be applied to other images and regions. Thus, the strength of the
optical signal coming from the various types of the seafloor is strongly influenced by the effects of the
water column, its depth and chemical properties. Yet methods of the unsupervised classification
overleap mixed spectral effects of the water column which shift the real values of the spectra, with the
pure reflectance from the benthos; as a result, it may cause significant errors \cite{Dierssen03}\label{Dierssen03}.

@@

Therefore, for the accurate assessments of various seafloor cover types, water
column depth and its optical properties, methods of supervised classification should be preferably
used as being more suitable in classification and interpretations of imagery.
While this and other studies \cite{Phinn08}\label{Phinn08}demonstrated the advantages and success of the
application of multispectral imagery for the spectral discrimination of the seafloor cover types and
mapping the submerged landscapes on the basis of pixels classification, the application of data from
hyperspectral sensors has better potential due to their higher resolution.

\subsubsection{Review of hyperspectral imagery used for seagrass mapping}
The application of the hyperspectral sensors is most effective and provide more accurate classification
results (Fig.2.1.), due to their higher spectral resolution \cite{Bharathi03}\label{Bharathi03} with interval
narrows to 10 nanometers, while broadband sensors are limited to the spectral width of ca
150 nm (Fig.\ref{fig:14}). 
Hyperspectral imagery is acquired through the simultaneous acquisition of images in many
narrow, contiguous spectral bands from hyperspectral scanners (mostly) cover the 400-
to 2500-nm spectral bands \cite{Schmidt03}\label{Schmidt03}.
Perhaps, the most advantageous and general characteristics of hyperspectral imagery is its
high spectral resolution, desirable for the case of seagrass monitoring. 
The most suitable
scanner fit for detailed seagrass mapping would cover bands of 550-750 nm and have a spectrum
resolution of 5-15 nm \cite{Fyfe04}\label{Fyfe04}, which exactly characterizes typical airborne and
hyperspectral satellite scanners. Important features may be detected in the narrow
wavelengths of hyperspectral imagery, while this information can be lost in the broader
wavelengths of other sensors. With its 126 spectral bands, HyMap imagery enables to
distinguish features of interest, i.e. seagrass types \cite{Peneva07}\label{Peneva07}, which is the major
advantage of the of hyperspectral data for mapping landscapes of the seagrasses.
While comparing multispectral imagery with airborne hyperspectral, the last one showed higher
overall accuracies \cite{Phinn08}\label{Phinn08}. Several, regular and narrow (10 nm) spectral bands, specific for
hyperspectral imagery, are strong tool enabling to detect even slight and subtle differences in the
spectral reflectances between various seafloor types, e.g. different seagrass species at diverse depths,
algae, corals, dark-coloured sands or other types of sediments \cite{Hochberg03a}\label{Hochberg03a}. 
Therefore, there is great potential of the application of hyperspectral remote sensing imagery towards seagrass
mapping at species level, as long as they are distinguishable spectrally (Fig.\ref{fig:15}) which, for example, has been
tested in the case study of Australian marine ecosystems by Fyfe \cite{Fyfe04}\label{Fyfe04}. Application of various
classifications methods, inter alia maximum likelihood, minimum distance and means, towards
hyperspectral imagery \cite{Peneva08}\label{Peneva08}, combined with fieldwork measurements ensures accurate
mapping results with the maximum likelihood methos producing the best results.
Therefore, accurate mapping of the seagrass landscapes and other seafloor types using remote sensing
approaches requires application of high-spatial resolution (higher than 5 m) or hyperspectral imagery.
Comparative analysis of the application of hyperspectral and multispectral imagery towards the
seafloor types classification \cite{Hochberg03a}\label{Hochberg03a} demonstrates that coral, seagrasses and
sand wery vell distinguishable in their spectra with an overall classification accuracy of 98 percent.
However, the use of data from various sensors, both hyperspectral and multispectral, are possible and
reasonable, as soon as it meets the research specific objective. 
Thus, the use of multispectral imagery
with high spatial resolution is preferable to using hyperspectral medium resolution data in case of
mapping benthic vegetation in areas where the spatial heterogeneity is very high \cite{Vahtmae07}\label{Vahtmae07}.

@@

In section \nameref{sec:2} we briefly discussed main existing cartographic methods and tools for seagrss mapping. In current  research we try to combine various techniques for assessment of seagrass in Mediterranean environment, in the cse study of Crete. Therefore, logically separated research sections are based on various software and tools. thus, spectral analysis and assessment of spectral signaturesis technically based on WASI, sttistical analysis of series of experimental raw data has been made by means of statistical software, such as \href{http://www.spss.com/}{SPSS}; mapping has been made using  \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS}; images processing and classification were made using\href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine} software by means of Google Earth and  \href{http://landsat.gsfc.nasa.gov/}{Landsat} imagery.
\href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0} software is used for the spatial analysis, general mapping and cartographic layout
presentation; the raster processing techniques are applied for the detection of seagrass spatial
distribution using supervised classification.
The research data include Google Earth aerial images and scenes from the \href{http://landsat.gsfc.nasa.gov/}{Landsat TM and ETM+}
covering research period of 10 years in the same year time, taken from \href{http://www.usgs.gov/pubprod/}{USGS}, \href{http://glovis.usgs.gov/}{GloVis}, \href{http://www.osdpd.noaa.gov/ml/index.html}{NOAA} and \href{http://glcfapp.glcf.umd.edu:8080/esdi/index.jsp}{The Earth Science Data Interface}. 
The satellite imagery provides
vital information of the most recent changes in \textit{P.oceanica} within the coastal areas, as well as the
condition (poor or destroyed).
The raster processing includes making mosaic-like covering for the whole research area. The Google
Earth images are most appropriate for the detailed mapping than stellite than \href{http://landsat.gsfc.nasa.gov/}{Landsat} or \href{http://asterweb.jpl.nasa.gov/}{ASTER} enabling to produce accurate maps with correct results. Therefore, the main set of images for the
current work is Google Earth aerial images. The  \href{http://landsat.gsfc.nasa.gov/}{Landsat} images are used for the general overview.
Besides aerial and satellite photographs, data acquired during the fieldwork (three weeks in September - October) are
necessary addition to the mapping helping to solve problems of interpretation during the images
classification \cite{Pasqualini98a}\label{Pasqualini98a}. Therefore, this work includes sampling of the in-situ
measurements of the seagrass distribution.
The sampling stations were located in two candidate places on the northern (Ligaria) and southern
(Xerocampus) parts of Crete island, as these regions are well suitable for the seagrass, due to the
annual mean water temperatures and geological factors, i.e. seafloor conditions and sediments. The
field campaign has been carried out during the September-October period 2010. 
The information about the location of the seagrass (mostly represented by the P.oceanica species) is useful for the
understanding of the relationship between the spatial distribution of the seagrass and the environment
of the selected areas of Cretan beaches. The results of the videographic measurements are used for the
seafloor types detection, because the objects represented on the photos can be well distinguished and
classified according to the following well-known characteristics \cite{Butler87}\label{Butler87}: size (yet some
measurements are necessary for the similar-looking objects); shape (the general form is most reliable
evidence for identification); colour (common and reliable object indicator); texture (when changes in
tone are too small to be distinguishable, texture may assist identification, e.g., stippled, granular,
rough, smooth, etc.); associated features (those usually found near other objects, e.g., rocks and soil).

@@

The Ligaria Beach is a narrow, sandy and pebble beach (Fig.12), located ca 15 km north-west from Heraklion.
The in-situ measurements were conducted during the fieldwork in the period 21.09.2010-11.10.2010.
The fieldwork on seagrass monitoring included visual estimations and photo- and video footage of the
above-ground seagrass patches, sediment seafloor cover types, species compositions, water depth and
geograpic locations recorded using GPS.
\pagebreak

@@

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras Olympus ST 8000 (Fig.\ref{fig:18}), at
five minutes interval along the tracklog path (Fig.\ref{fig:21}).
research sampling design also included measurements of depths (Fig. \ref{fig:63}), because bathymetry is one of the
most determinating factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, iPAQs, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The GPS allowed to capture measurement locations on the iPAQ,
encapsulated in a plastic waterproof Otterbox (Fig \ref{}). Camera were adjusted horizontally by a leveller and
were mounted under the bottom of the boat to capture video footage and imagery (Fig.\ref{fig:20}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. The
locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 m2 size each (Fig \ref{66}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects.

@@

Along the Ligaria beach there are several types of the seagrass landscapes, namely, segrass meadows
continuously covering vast areas (Fig.\ref{fig:62} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:62} (b)) and isolated seagrass patches, or mattes,
which located separately from broad meadows (Fig.\ref{fig:62} (c)).
The results of the underwater videometric measurements of the Olympus cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass P.oceanica meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig. \ref{fig:65}). Seagrass species on the Ligaria and Agia
Pelagia beach consist of P.oceanica. The types of sediment on the Ligaria vary from coarse sand,
P.oceanica patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:64}).

@@

\subsection{Review of the collected data}
The remote sensing imagery were collected for the Crete island area covering the research period and
area, enabling the field data to be used for the calibration and validation. The available aerial and
satellite imagery are commonly used for the mapping of seagrass landscapes and their application is
proven by various research papers. The imagery include satellite multi-spectral imagery (Landsat-TM,
ETM+) and aerial imagery from the Google Earth.
The overview of the collected data enables to summarize their following types:
\begin{itemize}
\item[] 1. Optical spectra of P.oceanica, carbonate sand, seawater with sediments and seawater measured in aquarium tank, without sediments, at different environmental conditions (e.g., Fig. \ref{fig:23})
\item[]2. Aerial imagery from the \href{http://www.google.com/earth/index.html}{Google Earth}
\item[]3. Satellite images from various open sources (e.g. Landsat, Tab \ref{tab:20}) (previews: Fig. \ref{fig:68})
\item[]4. Results of underwater videometric measurements of the Olympus cameras made during the ship route
\end{itemize}
The available satellite imagery (Fig.\ref{fig:68}) and aerial images are read into the ArcGIS project.
The available broadband and hyperspectral remote sensing data are used for the mapping of the
seagrass in shelf areas no deeper than 4.0 meters, for the environmental monitoring in order to detect
the dynamics in changes of the seagrass distribution along Crete during the past 10 years using
different satellite images for 2000-2010.
The spatial resolution of Landsat ETM+ image is 30 m in the visible and near infrared bands (bands
1-5 and 7); the spatial resolution of ASTER 15 m for the visible and near-infrared bands. IKONOS
acquires data in 3 visible channels and NIR, with spatial resolution of 1-4 meters. ASTER and
IKONOS images are suggested to be included as soon as available, in addition to the Landsat. The
research includes fieldwork in-situ measurements of the seagrass distribution along the northern and
south-eastern coasts of Crete in chosen locations.\\
The optical measurements of the irradiance and radiance of the sea water and bottom cover types of
the seafloor have been received by the means of the optical sensors Trios-RAMSES Hyperspectral
UVA/UVB Irradiance Sensor and RAMSES-ARC Hyperspectral UV-VIS Radiance Sensor (Fig. \ref{fig:10}), both adjusted for the measurements of the irradiance and radiance (see Appendices: Tab. \ref{tab:18}, Tab. \ref{tab:19}).
The highest measured values are located in the diapason of 410-730 nm for the water irradiance (Fig. \ref{fig:49}) and 430-650 for the water radiance (Fig.\ref{fig:47}). 
Afterwards, the measured values of the
radiance and irradiance, respectively, have been used for the computation of the spectral reflectance
properties of the sea water and bottom cover types (Fig.\ref{fig:23}). The spectral range of radiance
cover diapason of 320-950 nm, and ittadiance measurements are covered in the interval of 280-500
nm, which is suitable for a characteristics of seagrass reflectance.
Different curves on the reflectance, radiance and irradiance graphs for example, on Fig.\ref{fig:43}, or \ref{fig:45}, enlarged) represent several series of
the measurements. The values of the spectral reflectance are received from the
computations of these values using Excel mathematical formulae. The graphs shown on Fig.\ref{fig:47}, Fig. \ref{fig:48}, Fig. \ref{fig:49} display values of the radiance and irradiance of the sea water in Agia Pelagia, with and without sediments and suspended particles, respectively. Graph on Fig \ref{fig:44} displays statistuical analysis of the measured sets of observations, i.e. shows the midspread of the statistical quartiles (Q1-Q3), mean and extreme values.
\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{GNU-17.jpg}
\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete}
\label{fig:23}
\end{figure}
